Optimizing the future of biodiversity sampling by citizen scientists

### 2019-04-17 paper/main.tex
minor changes
updated figures

\begin{abstract}
Monitoring biodiversity trends in space and time is critical for conservation, but funding for ecology and conservation research is disappearing. Scientists are increasingly turning to citizen science data to extract information about population trends, with the aims of providing management recommendations. But these data are generally 'noisy' with a suite of biases that need to be statistically accounted for while estimating population trends. Undoubtedly, the immense increase in citizen science projects, and data, will continue to form an important resource for monitoring biodiversity in the future. Here, we ask whether these data can be improved --- maximizing the information content extracted from citizen science datasets --- by optimizing sampling in space and time. We provide a novel framework, based on statistical leverage, which assigns every citizen science observation a marginal value derived from the end statistical goal: in this case, population trends. We test how a suite of dynamic characteristics quantifying sampling in space and time influence the marginal value of an observation. We find that for a given grid, the days since the last sample and the distance from the nearest sampled grid are positively associated with the marginal value of a citizen science observation. Our framework is also independent to spatial-scale, easily adapted within a National Park, or an entire state or country. We conclude by using our parameterized model to produce dynamic maps, in space and time, predicting the expected marginal value in space for every day in 2018. Ultimately, we envision an approach where citizen scientists are guided and incentivized to sample sites with the highest expected marginal value.
\end{abstract}

@@

\begin{figure}[!hb]
\centering
\includegraphics[width=.8\linewidth]{param_estimates.png}
\caption{Six different days in 2018, showing a map of the Greater Sydney Region, and the predicted expected marginal value (i.e., statistical leverage) of biodiversity sampling that day. We predicted the expected value of citizen science observations for every day in 2018 (Video S2).}
\label{fig3}
\end{figure}

>>>>

\begin{abstract}
We are currently in the midst of the sixth extinction, and conservation biology and ecology are increasingly interested in tracking biodiversity trends in space and time. Yet, funding for ecology and conservation research is simultaneously disappearing. In turn, scientists are increasingly relying on citizen science data to extract information about population trends, with the aims of providing management recommendations. But these data are generally 'noisy' with a suite of biases that need to be statistically accounted for while estimating population trends. Undoubtedly, the immense increase in citizen science projects, and data, will continue to form an important resource for monitoring biodiversity in the future. Here, we ask whether these data can be improved --- maximizing the information content extracted from citizen science datasets --- by optimizing sampling in space and time. We provide a novel framework, based on statistical leverage, which assigns every citizen science observation a marginal value derived from the end statistical goal: in this case, population trends. We test how a suite of dynamic characteristics quantifying sampling in space and time influence the marginal value of an observation. We find that for a given grid, the days since the last sample and the distance from the nearest sampled grid are positively associated with the marginal value of a citizen science observation. Our framework is also independent to spatial-scale, easily adapted within a National Park, or an entire state or country. We conclude by using our parameterized model to produce dynamic maps, in space and time, predicting the expected marginal value in space for every day in 2018. Ultimately, we envision an approach where citizen scientists are guided and incentivized to sample sites with the highest expected marginal value.
\end{abstract}

@@

\begin{figure}[!hb]
\centering
\includegraphics[width=.8\linewidth]{example_maps.png}
\caption{Six different days in 2018, showing a map of the Greater Sydney Region, and the predicted expected marginal value (i.e., statistical leverage) of biodiversity sampling that day. We predicted the expected value of citizen science observations for every day in 2018 (Video S2).}
\label{fig3}
\end{figure}
